Mixture Models for Single-Cell Assays with Application to Vaccine Studies

### 2012-07-30 MIMOSA.tex
Make some more changes to the manuscript.

\begin{abstract}
Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on single-cell assays to provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential marker expression using single-cell assays. Our model allows the inference to be individual specific, while borrowing strength across individuals through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
\end{abstract}

@@

During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations. 
For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others thresholds are given by the assay technology. 
This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
 After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers. 
Using this matrix, one can form $2^K$ putative cell subsets. 
When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. 
A common statistical problem is, for a given marker combination, to identify individuals for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify individuals' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. 
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. 
 These antigen-specific T-cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells.

The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine. 
Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. 
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted. 
These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences. Individuals who generate a response after stimulation are called \textit{responders} whereas individual that do not show any differences are called \textit{non-responders}. 
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.

Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}. 
All of these methods test individuals separately, and no information is shared across individuals\footnote{We should say at some point that here we use individuals but the approach is general, could be used with animals, for example} even though one could expect some similarities across responders (\textit{resp.} non-responders). %In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here? Let's update this sentence later}.

The framework developed in this paper addresses these issues explicitly. In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across individuals by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood. 
In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across individuals. 
This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders. 
Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}. Our paper is organized as follows, ...

@@

\textit{Flow cytometry}:
The primary ICS data set is from a  trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen with 120 individuals (98 vaccines and 20 placebo recipients, see Web Appendix A). 
The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
Here, we analyze a subset of the data consisting of 98 individual from the vaccine group at two time points, day 0 and day 182. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. 
Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many individuals are expected to respond, for some cytokines at least. More details can be found in Web Appendix A.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen individuals. T-cells isolated by flow cytometry from sixteen individuals were stimulated in blocks of four individuals with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from the simulated samples  was compared to paired, unstimulated controls~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al. could you fill this in?}.

In the remainder of the paper, we use the following notation to describe our model.  From this point on, we assume that we observe cell counts from $I$ individuals in two conditions: stimulated and un-stimulated. Each cell can either be positive or negative for a marker. Given a set of $K$ markers, the measured cells can be classified into $2^K$ positive/negative marker combinations. We denote by $n_{sik}$ and $n_{uik}$, $k=1\dots 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples. We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for individual $i$ in each sample. For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for individual $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$. Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.

\section{Differential expression with one marker}
\label{s:DEone}
Datasets like the ones presented here are usually analyzed one marker at a times to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations. As a consequence, we first consider the one marker case (\textit{i.e.} $K=1$) where marginal cell counts are analyzed separately. In this case, for a given individual, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

\begin{table}
\centering
\parbox{0.8\linewidth}{
\caption{2 x 2 contingency table of counts for marker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given individual $i$.}\label{tab:twobytwo}
\centering
\begin{tabular}{rrr}

@@

For a given individual and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample. Individuals that show differential expression will be called responders for that marker. In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described in what follows.

\subsection{Beta-binomial model}
\label{s:DE}
For a given individual $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,

\begin{equation*}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation*}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding individuals we consider two competing models:
\begin{equation*}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:models}
\end{equation*}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal. Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder. In some study, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$. This alternative parametrization is described in Web Appendix C and we refer to it as the one-sided model.

\subsection{Priors}
\label{ss:priors}
Our model shares information across all individuals using exchangeable Beta priors on the unknown proportions, as follows, 
 \begin{align*}
(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\\
(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u)
 \end{align*}
where $z_i$ is an indicator variable equal to one if individual $i$ is a responder, i.e.  ${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all individuals. Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models, whereas $\alpha_s,\beta_s$ are only present in the alternative model. Finally, we assume that the $z_i$'s are independent and identically distributed Bernoulli with probability $w$, where $w$ represents the (unknown) proportion of responders. It follows that marginally, \textit{i.e.} after integrating $z_i$, $p_{ui}$ and $p_{si}$ are jointly distributed as a mixture of a one dimensional and a two dimensional Beta distributions with mixing parameter $w$. Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using and Expectation-Maximization~\citep{Dempster:1977ul} algorithm as described in Section \ref{s:estimation}. As an alternative, we also describe a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1. In this case, all parameters will be estimated via a Markov Chain Monte Carlo algorithm as described in Section \ref{s:estimation}.

@@

\subsection{EM algoritm}
Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
\[
\begin{split}
&\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\\ &\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
\end{split}
\] 
The M-step then consist of optimizing the complete-data log-likelihood over $\boldsymbol{\theta}$ after replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}. Straightforward calculations lead to 
$\tilde w = \sum_i{\tilde{z_i}}/I$, but unfortunately no closed form solutions exist for the remaining parameters. We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters.  Starting from some initial values, the EM algorithm iterates between the E and M steps until convergence. In our case, we initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative model components. We then use the estimated $z_i$'s to estimate the $p_{ui}$'s and $p_{si}$'s and use these to set the hyper-parameters to their method-of-moments estimates.

\subsection{MCMC algorithm}
Realizations were generated from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
Details about the algorithms are given in Web Appendix B. We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler. For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.

\section{Results}
\label{s:results}
In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data. The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate), defined as $\frac{\text{\# of false positives}}{\text{\# of observations called positive}}$ against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}.

@@

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+? Didn't look at CD8's}. The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A. 
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065} B). These results are consistent for other cytokines (see Web Figure A).

@@

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (see Web Figure B panels C and D). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.

>>>>

\begin{abstract}
Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on single-cell assays to provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential marker expression using single-cell assays. Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
\end{abstract}

@@

During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations. 
For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others thresholds are given by the assay technology. 
This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
 After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers. 
Using this matrix, one can form $2^K$ putative cell subsets. 
When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. 
A common statistical problem is, for a given marker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. 
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future~\citep{McKinstry:2010ei}. These antigen-specific T-cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells. The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine. 
Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. 
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted. 
These counts are compared between antigen stimulated and unstimulated samples from an subject to identify significant differences. Subjects who generate a response after stimulation are called \textit{responders} whereas subject that do not show any differences are called \textit{non-responders}. 
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.

Although there is no standard approach to analyzing single-cell assays, current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}\footnote{Can we match the citations with the methods used}. 
All of these methods test subjects separately, and no information is shared across subjects\footnote{We should say at some point that here we use subject but the approach is general, could be used with animals, for example} even though one could expect some similarities across responders (\textit{resp.} non-responders). %In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here? Let's update this sentence later}.

The framework developed in this paper addresses these issues explicitly. In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across subjects by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood. 
In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across subjects. 
This sharing of information across subjects helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders. 
Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}. Our paper is organized as follows, Section \ref{s:data} introduces the data and notations used in the paper. In Section \ref{s:DEone}, we present our model for testing differential marker expression in the univariate case. Section \ref{s:results} compares our approach to alternative methods and test the robustness of our model. In Section \ref{s:demarkercombos} we present a multvariate extension of our model that can be used to test multivariate marker differential expression and present some results using our single cell gene expression data. Finally, in Section \ref{s:discussion} we discuss our findings and future work.

@@

\textit{Flow cytometry}:
The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA\footnote{Define} vaccines in a prime-boost regimen with 120 subjects \citep{Goonetilleke:2006jk}. 
The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
Here, we analyze a subset of the data consisting of 98 subject from the vaccine group at two time points, day 0 and day 182. 
Three cytokines (IFN$\gamma$, TNF$\alpha$ and IL2) were measured at the single-cell level for each subject and time point, with and without stimulations with an antigen (HIV Envelope peptide pool) matching part of the vaccine. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen subjects. T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from the simulated samples  was compared to paired, unstimulated controls~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al. could you fill this in?}.

In the remainder of the paper, we use the following notation to describe our model.  From this point on, we assume that we observe cell counts from $I$ subjects in two conditions: stimulated and un-stimulated. Each cell can either be positive or negative for a marker. Given a set of $K$ markers, the measured cells can be classified into $2^K$ positive/negative marker combinations. We denote by $n_{sik}$ and $n_{uik}$, $k=1\dots 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples. We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for subject $i$ in each sample. For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$. Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.

\section{Differential expression with one marker}
\label{s:DEone}
Datasets like the ones presented here are usually analyzed in in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations. Here, by univariate, we mean that we have only one positive cell subset. This cell subset can be defined as considering by the expression of one marker alone (marginalizing over all other measured markers) such as IL2$+$ (\textit{vs.} IL2$-$), or considering a specific positive marker combination (and marginalizing over everything else) such as IL2$+$ and/or INF$\gamma+$ (\textit{vs.} IL2$-$/INF$\gamma-$). Without loss of generality, we treat the univariate case as a one marker case (\textit{i.e.} $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

\begin{table}
\centering
\parbox{0.8\linewidth}{
\caption{2 x 2 contingency table of counts for marker positive and negative cells between stimulated ($s$) and unstimulated ($u$) conditions for a given subject $i$.}\label{tab:twobytwo}
\centering
\begin{tabular}{rrr}

@@

For a given subject and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample. Subjects that show differential expression will be called responders for that marker. In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described in what follows.

\subsection{Beta-binomial model}
\label{s:DE}
For a given subject $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,

\begin{equation*}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation*}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding subjects we consider two competing models:
\begin{equation*}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:models}
\end{equation*}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal. Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder. In some study, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$. This alternative parametrization is described in Web Appendix C and we refer to it as the one-sided model.

\subsection{Priors}
\label{ss:priors}
Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions, as follows, 
 \begin{align*}
(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\\
(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u)
 \end{align*}
where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, i.e.  ${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all subjects. Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models, whereas $\alpha_s,\beta_s$ are only present in the alternative model. Finally, we assume that the $z_i$'s are independent and identically distributed Bernoulli with probability $w$, where $w$ represents the (unknown) proportion of responders. It follows that marginally, \textit{i.e.} after integrating $z_i$, $p_{ui}$ and $p_{si}$ are jointly distributed as a mixture of a one dimensional and a two dimensional Beta distributions with mixing parameter $w$. Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using and Expectation-Maximization~\citep{Dempster:1977ul} algorithm as described in Section \ref{s:estimation}. As an alternative, we also describe a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1. In this case, all parameters will be estimated via a Markov Chain Monte Carlo algorithm as described in Section \ref{s:estimation}.

@@

\subsection{EM algoritm}
Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
\[
\begin{split}
&\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\\ &\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
\end{split}
\] 
The M-step then consist of optimizing the complete-data log-likelihood over $\boldsymbol{\theta}$ after replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}. Straightforward calculations lead to 
$\tilde w = \sum_i{\tilde{z_i}}/I$, but unfortunately no closed form solutions exist for the remaining parameters. We use numerical optimization as implemented in R\footnote{Add citation to R}'s \textit{optim} function to estimate the remaining parameters.  Starting from some initial values, the EM algorithm iterates between the E and M steps until convergence. In our case, we initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative model components. We then use the estimated $z_i$'s to estimate the $p_{ui}$'s and $p_{si}$'s and use these to set the hyper-parameters to their method-of-moments estimates.

\subsection{MCMC algorithm}
Realizations were generated from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
Details about the algorithms are given in Web Appendix B. We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler. For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities\footnote{Need to check this}. Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.

\section{Results}
\label{s:results}
In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data. The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Storey:2002vj}.

@@

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A. 
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positive being detected. Our comparison also show that ranking based on log-fold change alone is not reliable and should not be used.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065} B). Here we presented the results based on IL2 and IFN$\gamma$ alone and the subset IL2 and/or IFN$\gamma$ that were used in the original study\footnote{reference}. These results are consistent for other cytokines and cytokine combinations (see Web Figure A).

@@

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the ICS data set (see Web Figure B panels C and D). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.
